---
title: "Reproducible Research Course Project 1"
author: "Gbemisola Oni"
date: "19 July 2018"
output: html_document
---

##***Loading and preprocessing the data***
```{r loaddata, echo=TRUE}
#read data
activitydata <- read.csv("activity.csv", header = TRUE, sep = ",")
head(activitydata)
dim(activitydata)
```

##***What is mean total number of steps taken per day?***

###1.Calculate the total number of steps taken per day.
```{r meansteps, echo=TRUE}
# Change date type to date class 
activitydata$date <- as.Date(as.character(activitydata$date))
activitydataNA <- is.na(activitydata$steps)
# Create data without NAs
activitydatatidy <- activitydata[!activitydataNA,]

# Aggregate clean non NA steps per day (SUM)
TotalDataByDay <- aggregate(activitydata$steps, by=list(activitydata$date), sum)
# Create column names
names(TotalDataByDay)[1] ="Date"
names(TotalDataByDay)[2] ="Steps"
head(TotalDataByDay,10)
#dimensions of full dataset
dim(TotalDataByDay)
```

###2. Make a histogram of the total number of steps taken each day.
```{r plothist, echo=TRUE}
library(ggplot2)

ggplot(TotalDataByDay, aes(x=Steps)) +geom_histogram(fill="deeppink3")+labs(title="Total Steps per Day", x="Steps", y="Frequency")

```

###3.Calculate and report the mean and median of the total number of steps taken per day

```{r mean_median, echo=TRUE}
mean(TotalDataByDay$Steps, na.rm = TRUE)

median(TotalDataByDay$Steps, na.rm = TRUE)
```

##***What is the average daily activity pattern?***

###1. Make a time series plot (i.e.type="l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r time_series, echo=TRUE}
#create subset without NAs
activitydatatidy <- activitydata[!activitydataNA,]
meanDataByInterval <- aggregate(activitydatatidy$steps,by=list(activitydatatidy$interval), mean)
# Create column names
names(meanDataByInterval)[1] ="interval"
names(meanDataByInterval)[2] ="steps"

ggplot(meanDataByInterval, aes(x = interval, y=steps)) +
  labs(title = "Total Steps by Interval", x = "Interval", y = "Steps") + geom_line(colour="deeppink3",size=2)
```

###2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r max_mean, echo=TRUE}
maxStepsInterval <- meanDataByInterval[which.max(meanDataByInterval$steps),]
maxStepsInterval
```

##***Imputing missing values***

###1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r missingvals, echo=TRUE}
countNaRows<- nrow(activitydata)- nrow(activitydatatidy)
countNaRows
```

###2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.
```{r fillNAS, echo=TRUE}
library(magrittr)
library(dplyr)

replaceNas <- function(x) replace(x, is.na(x), mean(x, na.rm = TRUE))
activitydata2 <- activitydata%>% group_by(interval) %>% mutate(steps= replaceNas(steps))
#show that there are no NA rows

sum(is.na(activitydata2$steps))

```

###3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r datasetfilledNAS, echo=TRUE}
#show first 6 rows of dataset created in previous section
head(activitydata2)
```

###4a. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. 
```{r plothistogram, echo=TRUE}
TotalDataByDay2 <- aggregate(activitydata2$steps, by=list(activitydata2$date), sum)

names(TotalDataByDay2)[1] ="Date"
names(TotalDataByDay2)[2] ="Steps"
head(TotalDataByDay2,10)
#dimensions of full dataset 
dim(TotalDataByDay2)

#plot histogram
ggplot(TotalDataByDay2, aes(x=Steps)) +geom_histogram(fill="deeppink3")+labs(title="Total Steps per Day", x="Steps", y="Frequency")

#mean
mean(TotalDataByDay2$Steps, na.rm = TRUE)

#median
median(TotalDataByDay2$Steps, na.rm = TRUE)

```

###4b. Do these values differ from the estimates from the first part of the assignment? 

The mean is the same but the median is slightly higher than when there were missing values.Compare results below.

```{r compare, echo=TRUE}
#compare mean before and after removing NA values
mean(TotalDataByDay$Steps, na.rm = TRUE)

mean(TotalDataByDay2$Steps)

#compare median before and after removing NA values
median(TotalDataByDay$Steps,na.rm =TRUE)

median(TotalDataByDay2$Steps)
```

###4c. What is the impact of imputing missing data on the estimates of the total daily number of steps?

The median of the dataset is now equal to the mean.

##***Are there differences in activity patterns between weekdays and weekends?***

###1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
```{r weekday_weekend, echo=TRUE}
#assigning weekday/ weekend tag to the 5 minute interval dataset
activitydata2$weekday <- weekdays(activitydata2$date)
activitydata2$weekend <- ifelse (activitydata2$weekday == "Saturday" | activitydata2$weekday == "Sunday", "Weekend", "Weekday")

head(activitydata2,15)

#assigning weekday/weekend tag to the daily interval set
TotalDataByDay2$weekday <- weekdays(TotalDataByDay2$Date)
TotalDataByDay2$weekend <- ifelse (TotalDataByDay2$weekday == "Saturday" | TotalDataByDay2$weekday == "Sunday", "Weekend", "Weekday")

head(TotalDataByDay2,15)
```


###2. Make a panel plot containing a time series plot (i.e. type="l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 

```{r panelplot, echo=TRUE}
#create Weekend and weekday subset dataframe
data.weekday <- activitydata2[activitydata2$weekend=="Weekday",]
data.weekend <- activitydata2[activitydata2$weekend=="Weekend",]

#Calculate mean of steps for 5 mins interval
weekdayinterval <- group_by(data.weekday, interval)
weekdayinterval <- summarize(weekdayinterval, steps=mean(steps))
weekdayinterval$weekend <- "Weekday"
weekendinterval <- group_by(data.weekend, interval)
weekendinterval <- summarize(weekendinterval, steps=mean(steps))
weekendinterval$weekend <- "Weekend"

#plot
activityinterval <- rbind(weekdayinterval, weekendinterval)
activityinterval$weekend <- as.factor(activityinterval$weekend)
ggplot(activityinterval, aes(interval, steps, color=weekend)) + geom_line() + facet_grid(weekend ~ .)
```

###At the beginning of the week, there is peak which drops and stays low . Over the weekend, there is a more uniform pattern.###





